First paper - Draft

### 2017-06-21 paper/paper.Rnw
Included about 80% of equations - haven't tested yet

\citet{reardon2011} finds that the increase has occured predominantely from the 1970's until the 2000's. In fact, hard numbers suggest that the gap increased by about ~ 40-50\%. The author also estimates the rate of change using data from the 1940's and finds an even higher increase of about ~ 75\%. Since the studies before the 1970's lack a lot of rigor in comparability and sampling, the author computes everything for the sample before/after 1970's separately \footnote{Even after weighting appropriately.}. Finally, \citet{readon2011} concludes that the U.S gap between the 90th and 10th income percentile is at about 1.25 standard deviations.  \citet{bradbury2015} don't find results which are very far away from this. They find that for 14 year olds in the United States, the gap is above 1 SD but lower than 1.25. \citet{duncan2011} find similarly to the previous studies and confirm a gap of ~ 1.25/1.50 standard deviations.

@@

Their work suggests that there is considerable variation in the achievement gap between top and bottom earning families across many developed countries. In comparison to the literature on achievement gaps, they find that the U.S has a gap a little over 1 standard deviation in the year 2005 (for 15 year olds) and Germany has a decreasing gap from around ~ 1.25 to ~ 1 standard deviation. They go even further and link this achievement gap to several country-level indicators related to income inequality, school differentiation, central exams, etc ... The correlations are indeed very suggestive but clearly we must be extremely cautios in drawing causal changes from these two variables. But it is important to stress that their design is quite different from the work of Reardon (2011). Reardon (2011) takes studies in the U.S starting from the 1940's until 2015 \footnote{Each study is independent of each other meaning that the it might've been given to 6 year olds as well as to 12 year olds. Although the author adjusts for age, the trends can't be generalized to age-specific, but rather in overall terms.} and makes them comparable across time. This gives the author a very long time series to build a reliable achievement gap (over 40 years). In their study, Chmielewski and Reardon (2016) change the aim of the study to model between-country differences from a cross-sectional perspective.

Chmielewski (2017), building on the work of Chmielewski and Reardon (2016) and Reardon and Portilla (2015) pooled together all the previously mentioned data, together with over 10 more studies ranging from the year 1964 until 2015 in order to discover differences between and across countries. With over 50 years of data, and over 100 countries, Chmielewski (2017) finds that there seems to be a general world increase in the achievement gap. However, once she disentangles the relationship by country, she finds a reasonable amount of heterogeneity, with some countries seeing the achievement gap closing, others no change at all, while others record a steady increase. One clear limitation of their study (as well as Reardon (2011)) is that the adjust for the age of each child in all studies. Although for their purposes is the right thing to do \footnote{The differences in achievement could simply be due to changes in cognitive abilities across the lifetime. However, as we've noted before, Corak and Waldfogel (2015) find that the achievement gap is very stable across the life time}, they are masking age-specific achievement gaps by controlling for age, such as Reardon (2011) did.

@@

\begin{description}
  \item[We want to test the notion that better performing countries have lower leves of achievement gap. \citep{werfhorst_mijs} emphasize that there is empirical evidence that suggests this. This pattern is not so obvious. For example, countries with high leves of tracking could maximize student performance, specially the high SES students, raising their overall performance and thus raising the national performance score. But if the bottom performers are not gaining at the same rate, then the achievement gap will inevitable grow resulting in a high performing countries with widening achievement gap.]
  \item[The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker that we though. After recording a SES gap increase of about 40\% in only 30 years, \citep{reardon_portilla} stress that they also found a significant decrease in only 15 years of data. This lends credibility that our estimations will show variation over time. As usual, their estimations are only for children going deom 4th to 8th grade. We concentrate on age-specific groups, in this case, 15 year olds.]
  \item[We want to compare the percentage change at which the gap widened/narrowed from the first to the last year. This will give us a general idea of the overall change over time, and will allow us to compare our estimates to the actual literature \footnote{Although no study has performed this age-specific achievement gap for comparable tests over such a long time. Our results will serve as comparison for other studies that use age-specific groups, such as 4th graders.}]
  \item[The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment. The literature has concentrated very narrowly on whether the gap is increasing because the top performers are getting ahead, because both are distancing or because the bottom is falling behind. We shall pay particular attention to identifying the rate at which the top/bottom groups are evolving over time. This type of analysis is particularly useful in jotting down the mechanisms through which the gap is evolving. If the bottom performers are stagnated whereas the top is gaining ground, the efficiency-trade off dilemma of tracking might be much more credible than other explanations.]
\end{description}

@@

In order to measure family background PISA includes several variables. Firstly, the ask student's their parent's educational level. Scholars have considered this to be a reliable recall given that we expect fifteen year olds to know their parent's level of education (Reardon 2011). This question has been asked in every wave and holds a somewhat similar coding across time, although the first two waves have a different coding. Moreover, another limitation is the fact that parent's education is measured with the ISCED classification, something that has changed over time. For example, until PISA 2009, the ISCED 1997 classification was currently used, but the next wave already used the new ISCED 2011 classification. Both these classification schemes have equivalent look-up tables, yet this requires a detailed inspection of the codings.

@@

Reardon (2011) and subsequently (all reardon papers), used a different method developed by Reardon (2011) which we partially adopt adopt in this paper. SES achievement gaps are measured as the difference in standardized achievement between the 90th and 10th percentiles of the chosen SES variable.

% Transform all of this text into equations

We start by standardizing achievement for each country, year and subject. Because PISA does not provide a single achievement indicator, we take the median of all plausible values for each student \footnote{Since each plausible value is a random draw from a theoretical latent normal distribution of possible student achievement scores, the median should be precise in getting a central measure of the latent distribution.}, resulting in one single score. We then fit a linear model for each wave where we regress the single achievement score on the age variable weighted by the appropriate student sample weights. We take the residuals and divide them by the root mean square error of the model and divide that by one over the square root of the reliability of the survey for that year-subject (implement this all in formula). \footnote{Other proceedures multiply each country by their own reliability measure for each year-subject pair (Anna analysis of wordwide gap). Unfortunately, PISA 2000 did not provide any reliability measure separately for each country and PISA 2015 has to yet release their own at the moment of writing this paper. For this reason we implement the analysis following the original work of Reardon (2011)}

@@

Inspection of the data suggests that it's something we shouldn't be deeply concerned with. The standard deviation of each wave seems to be following a very similar pattern with an exception of the year 2000.

Another concern is that if the gaps at different waves are measured with tests that have different amounts of measurement error, then the amount of bias will not be the same in each measure of the gap. This can be very misleading and suggest erroneous intepretations regarding trends in the of the gaps over time \citet{reardon2011}. The PISA has tried to make sure the tests are comparable across waves (PISA 2012 technical report), but still we have to adjust for this imprecision.

In order to correct gap estimates for measurement error, I weight the student test scores by the P where P is the reliability of the test for each wave. Each PISA study provided a global reliability indicator which we use accordingly to each wave. This yields estimates of the true gaps, and eliminates any bias in the trend that may arise from differential reliability of the tests.

After constructing the adjusted test score measurement, we estimate the SES gap in a very simple fashion. For each country-wave, we estimate the weighted 90th and 10th percentile scores for the SES composite index, and then we generate a dummy of 1 for those above (including) the 90th percentile and 0 for those below (including) the 10th percentile. Finally, for each wave, we fit a mixed model with the adjusted test score as the dependent variable and the SES index as the only covariate. We allow both the intercept and the dummy variable to vary by country, and weight the model by the appropriate student weights.

>>>>

\citet{reardon2011} finds that the increase has occured predominantely from the 1970's until the 2000's. In fact, hard numbers suggest that the gap increased by about ~ 40-50\%. The author also estimates the rate of change using data from the 1940's and finds an even higher increase of about ~ 75\%. Since the studies before the 1970's lack a lot of rigor in comparability and sampling, the author computes everything for the sample before/after 1970's separately \footnote{Even after weighting appropriately.}. Finally, \citet{reardon2011} concludes that the U.S gap between the 90th and 10th income percentile is at about 1.25 standard deviations.  \citet{bradbury2015} don't find results which are very far away from this. They find that for 14 year olds in the United States, the gap is above 1 SD but lower than 1.25. \citet{duncan2011} find similarly to the previous studies and confirm a gap of ~ 1.25/1.50 standard deviations.

@@

Their work suggests that there is considerable variation in the achievement gap between top and bottom earning families across many developed countries. In comparison to the literature on achievement gaps, they find that the U.S has a gap a little over 1 standard deviation in the year 2005 (for 15 year olds) and Germany has a decreasing gap from around ~ 1.25 to ~ 1 standard deviation. They go even further and link this achievement gap to several country-level indicators related to income inequality, school differentiation, central exams, etc ... The correlations are indeed very suggestive but clearly we must be extremely cautios in drawing causal changes from these two variables. But it is important to stress that their design is quite different from the work of \citet{reardon2011}. \citet{reardon2011} takes studies in the U.S starting from the 1940's until 2015 \footnote{Each study is independent of each other meaning that the it might've been given to 6 year olds as well as to 12 year olds. Although the author adjusts for age, the trends can't be generalized to age-specific, but rather in overall terms.} and makes them comparable across time. This gives the author a very long time series to build a reliable achievement gap (over 40 years). In their study, Chmielewski and Reardon (2016) change the aim of the study to model between-country differences from a cross-sectional perspective.

Chmielewski (2017), building on the work of Chmielewski and Reardon (2016) and Reardon and Portilla (2015) pooled together all the previously mentioned data, together with over 10 more studies ranging from the year 1964 until 2015 in order to discover differences between and across countries. With over 50 years of data, and over 100 countries, Chmielewski (2017) finds that there seems to be a general world increase in the achievement gap. However, once she disentangles the relationship by country, she finds a reasonable amount of heterogeneity, with some countries seeing the achievement gap closing, others no change at all, while others record a steady increase. One clear limitation of their study (as well as \citet{reardon2011}) is that the adjust for the age of each child in all studies. Although for their purposes is the right thing to do \footnote{The differences in achievement could simply be due to changes in cognitive abilities across the lifetime. However, as we've noted before, Corak and Waldfogel (2015) find that the achievement gap is very stable across the life time}, they are masking age-specific achievement gaps by controlling for age, such as \citet{reardon2011} did.

@@

a) We want to test the notion that better performing countries have lower leves of achievement gap. \citep{werfhorst_mijs} emphasize that there is empirical evidence that suggests this. This pattern is not so obvious. For example, countries with high leves of tracking could maximize student performance, specially the high SES students, raising their overall performance and thus raising the national performance score. But if the bottom performers are not gaining at the same rate, then the achievement gap will inevitable grow resulting in a high performing countries with widening achievement gap.

b) The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker that we though. After recording a SES gap increase of about 40\% in only 30 years, \citep{reardon_portilla} stress that they also found a significant decrease in only 15 years of data. This lends credibility that our estimations will show variation over time. As usual, their estimations are only for children going from 4th to 8th grade. We concentrate on age-specific groups, in this case, 15 year olds.

c) We want to compare the percentage change at which the gap widened/narrowed from the first to the last year. This will give us a general idea of the overall change over time, and will allow us to compare our estimates to the actual literature \footnote{Although no study has performed this age-specific achievement gap for comparable tests over such a long time. Our results will serve as comparison for other studies that use age-specific groups, such as 4th graders.}

d) The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment. The literature has concentrated very narrowly on whether the gap is increasing because the top performers are getting ahead, because both are distancing or because the bottom is falling behind. We shall pay particular attention to identifying the rate at which the top/bottom groups are evolving over time. This type of analysis is particularly useful in jotting down the mechanisms through which the gap is evolving. If the bottom performers are stagnated whereas the top is gaining ground, the efficiency-trade off dilemma of tracking might be much more credible than other explanations.

@@

In order to measure family background PISA includes several variables. Firstly, the ask student's their parent's educational level. Scholars have considered this to be a reliable recall given that we expect fifteen year olds to know their parent's level of education \citep{reardon2011}. This question has been asked in every wave and holds a somewhat similar coding across time, although the first two waves have a different coding. Moreover, another limitation is the fact that parent's education is measured with the ISCED classification, something that has changed over time. For example, until PISA 2009, the ISCED 1997 classification was currently used, but the next wave already used the new ISCED 2011 classification. Both these classification schemes have equivalent look-up tables, yet this requires a detailed inspection of the codings.

@@

\citet{reardon2011} and subsequently (all reardon papers), used a different method developed by \citet{reardon2011} which we partially adopt adopt in this paper. SES achievement gaps are measured as the difference in standardized achievement between the 90th and 10th percentiles of the chosen SES variable.

We start by standardizing achievement for each country, year and subject. Because PISA does not provide a single achievement indicator, we take the median of all plausible values for each student \footnote{Since each plausible value is a random draw from a theoretical latent normal distribution of possible student achievement scores, the median should be precise in getting a central measure of the latent distribution.}, resulting in one single score.

We then fit a linear model

\begin{equation}
Y_i = \alpha + \beta_1 * AGE_i + \epsilon_i, \epsilon_i ~ N(0, \sigma^2)
\end{equation}

for each wave where \begin{math}Y_i\end{math} is the median student test score and \begin{math}AGE_i\end{math} is their age measured in months (because all children should be 15 year olds) weighted by the student sample weights.

We then adjust \begin{math} \hat{Y_i} \end{math} by

% I think the denominator could be wrong! Because I include y_hat and that's what the equation is estimating
\begin{equation}
\hat{Y_i} = \frac{\hat{\epsilon_i}}{\sqrt{1/n \sum_{i=1}^{n} (y_i - \hat{y_i})^2}}
\end{equation}

% # Take residuals of model and divide by rmse. Multiply that by
% # 1 / sqrt(reliability of each survey), which is .y in the loop.
% .x$adj_pvnum <- resid(mod1)/rmse(mod1, .x) * 1 / sqrt(.y)

Where \begin{math}\epsilon_i\end{math} is the residual for student \begin{math}i\end{math} and the denominator is the root mean square error of the model.

@@

Inspection of the data suggests that it's something we shouldn't be deeply concerned with. The standard deviation of each wave seems to be following a very similar pattern with not so drastic exception of the year 2000.

Another concern is that if the gaps at different waves are measured with tests that have different amounts of measurement error, then the amount of bias will not be the same in each measure of the gap. This can be very misleading and suggest erroneous intepretations regarding trends in the of the gaps over time \citet{reardon2011}. The PISA has tried to make sure the tests are comparable across waves (PISA 2012 technical report), but we still have to adjust for this imprecision.

In order to correct gap estimates for measurement error, I adjust student test scores by

\begin{equation}
\hat{Y_i} = \hat{Y_i} * \frac{1}{\sqrt{r}}
\end{equation}

where \begin{math}r\end{math} is the reliability score of the wave. \footnote{Other proceedures multiply each country by their own reliability measure for each year-subject pair (Anna analysis of wordwide gap). Unfortunately, PISA 2000 did not provide any reliability measure separately for each country and PISA 2015 has to yet release their own at the moment of writing this paper. For this reason we implement the analysis following the original work of \citet{reardon2011}}. Each PISA survey provides a reliability indicator which we use accordingly. This yields estimates of the true gaps, and eliminates any bias in the trend that may arise from differential reliability of the tests.

After constructing the adjusted test score measurement, we estimate the SES dummy by calculating the weighted 90th and 10th percentile scores for the SES composite index, and then generating a dummy of 1 for those above (including) the 90th percentile and 0 for those below (including) the 10th percentile.

We then fit

\begin{equation}
\hat{Y_i} = \alpha_{j[i]} + \beta_{j[i]} * SES_i + \epsilon_i, \rlap{\quad \text{for} \(i\) \text{= 1, 2, ...,} \(n\) \text{within country} \(j\)}
\end{equation}

% mod2 <-
%   lmer(adj_pvnum ~ escs_dummy + (1 + escs_dummy | country),
%        data = .x,
%        weights = .x[[weights_var]])

where \(SES_i\) is whether the student is in the top/bottom SES dummy and we allow both \(\alpha_i\) and \(\beta_i\) to vary by country \(j\). We implement this separately for each wave and weighting by the student sample weights.

Finally,

### 2017-06-21 paper/paper.tex
Included about 80% of equations - haven't tested yet

Recent research on educational inequality has found that differences in test peformance between High-SES and Low-SES kids has been growing quickly over the years. The literature on educational inequality has mainly concentrated on the United States (reardon here) but other international evidence is emerging with other similar findings for other count. The United States is usually the case study of interest, firstly, because it is the only country where cognitive testing is very widespread across surveys. And secondly, this trend has allowed to have testing records as early as 1940 (check reardon) until present day. Using this information, Reardon (2011) is the first to investigate the evolution of the cognitive gap and the results are very surprising. Not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black gap (cite study of white-black gap). The widening of the achievement gap has been happening in parallel to the growth of income inequality. Although very suggestive, it is hard to link both things causally.

Reardon (2011) finds that the increase has occured predominantely from the 1970's until the 2000's. In fact, hard numbers suggest that the gap increased by about ~ 40-50\%. The author also estimates the rate of change using data from the 1940's and finds an even higher increase of about ~ 75\%. Since the studies before the 1970's lack a lot of rigor in comparability and sampling, the author computes everything for the sample before/after 1970's separately \footnote{Even after weighting appropriately.}. Finally, Reardon (2011) concludes that the U.S gap between the 90th and 10th income percentile is at about 1.25 standard deviations. Corak and Waldfogel (2015) don't find results which are very far away from this. They find that for 14 year olds in the United States, the gap is above 1 SD but lower than 1.25. Duncan and Magnuson find similarly to the previous studies and confirms a gap of ~ 1.25/1.50 standard deviations.

One important drawback of these studies is that they don't present the uncertainty of this estimate. Not necessarily to gauge their statistical significance, but to simply asses how much we can trust their accuracy. Despite the recent surge in understanding achievement gaps as a function of social background, sociologists have already amassed a great deal of findings concerning inter-generational inequalities, one notable example being that it's very clear that large inequalities in parent's education tend to reproduce large inequalities in children's education (Breen \& Jonsson 2007). Virtually in all countries, be it developed or developing, there is inequality of opportunity. But there is considerable variation in the magnitud of this effect. For example, the Scandinavian countries, particularly Denmark and Sweden, prove to be very mobile countries (Gosta and his phd student, Breen and other people, Shavit first study).

With very recent data, Reardon and Portilla (2015) have unexpectedly uncovered a new finding: the reversal of the trend. From 1998 until 2010, it seems like the income achievement gap \footnote{The income achievement gap is the difference between the 90th and 10th income percentile in the test of interest} decreased at a rate of 0.01 and 0.008 standard deviations per year. In contrast to Reardon (2011), we find that in a 30-year span the gap was systematicaly increasing at a rate of 0.02, something not dramatically different from the previous estimates. The American situation is a very distinctive one relative to Europe. The reasons why we're seeing a reversal in the trend could be numerous and should be studied very close to context. For example, the American educational system lacks any formal curricular differentiation \emph{à la} European style. In contrast, Reardon and Portilla (2015) suggest that the reversal is likely due to the high increase of preschool enrollment. They suggest that in this same period (1998 - 2010) the income achievement gap in early schooling enrollment decreased substantially.

Some authors have taken this analysis to an international context in order to discover between-country trends. The comprehensive work of Corak and Waldfogel () performs a comparative analysis of Australia, United Kingdom, United States and Canada. Their research design is very distinctive in that they use longitudinal data from children as early as age 2 and study the evolution of the achievement gap up until age 14 \footnote{To the best of my knowledge this is the only study that uses panel data to study achievement gaps, let alone to do this between countries}. The core finding behind the book is that the American achievement gap is much wider than in any other comparison countries. As mentioned earlier, they find that once the achievement gap is present in early school entry, it doesn't seem to narrow very much over the life course. They find that the achievement gap is very stable all the way from early education to secondary school (although their studies cover up until age 14).

One limitation of their study is that they concentrate on countries which have a very particular educational structure, namely the fact that there is little formal stratification in terms of curricula. These four countries have no major jump in school selection, quite the opposite to the European average countries. A thorough review by Van der Werfhorst and Mijs (2010) sheds some light on the subject. First and foremost, they gather substantive evidence showing that countries which have a highly tracked curriculum tend to have high levels of inequality, measured in terms of achievement gaps. Hanushek and Woessman (tracking study), use the TIMSS and PISA surveys to gauge whether highly tracked countries do indeed increase inequality after students pass the age at first selection of tracking. Moreover, Mickleweight \& Schnepf (2007) using similar data but different design finds that countries which have a high level of tracking, are distinctively unequal in the difference between the top 95th and bottom 5th performers. In fact, the difference in test scores between these two groups is about 10 times higher then the average annual gain of a year of schooling.

@@

@@

The aim of this paper is to identify and disaggregate country trends in the achievement gap for several countries. To represent the SES gap, most of the literature on achievement gaps has concentrated on indicators such as parental education, parental occupational status, income achievement gaps and actual SES achievement gaps. The actual calculation of the achievement gap varies substantially and different strategies have been implemented. For example, Mickleweight & Funsche calculate the difference in achievement by crudely subtracting the gap between the 95th and 5th percentile of the mathematics distribution. Although in principle you should be able to capture some type of SES effect, theoretically, it should be much more accurate to difference out the mean score of, for example, parental education.
\section{Analysis}

>>>>

Recent research on educational inequality has found that differences in test peformance between High-SES and Low-SES kids has been growing quickly over the years. The literature on educational inequality has mainly concentrated on the United States \citep{reardon2011} but other international evidence is emerging with other similar findings for other count. The United States is usually the case study of interest, firstly, because it is the only country where cognitive testing is very widespread across surveys. And secondly, this trend has allowed to have testing records as early as 1940 (check reardon) until present day. Using this information, \citep{reardon2011} is the first to investigate the evolution of the cognitive gap and the results are very surprising. Not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black gap (cite study of white-black gap -waldfogel study). The widening of the achievement gap has been happening in parallel to the growth of income inequality. Although very suggestive, it is hard to link both things causally.

\citet{reardon2011} finds that the increase has occured predominantely from the 1970's until the 2000's. In fact, hard numbers suggest that the gap increased by about ~ 40-50\%. The author also estimates the rate of change using data from the 1940's and finds an even higher increase of about ~ 75\%. Since the studies before the 1970's lack a lot of rigor in comparability and sampling, the author computes everything for the sample before/after 1970's separately \footnote{Even after weighting appropriately.}. Finally, \citet{reardon2011} concludes that the U.S gap between the 90th and 10th income percentile is at about 1.25 standard deviations.  \citet{bradbury2015} don't find results which are very far away from this. They find that for 14 year olds in the United States, the gap is above 1 SD but lower than 1.25. \citet{duncan2011} find similarly to the previous studies and confirm a gap of ~ 1.25/1.50 standard deviations.

One important drawback of these studies is that they don't present the uncertainty of this estimate. Not necessarily to gauge their statistical significance, but to simply asses how much we can trust their accuracy. Despite the recent surge in understanding achievement gaps as a function of social background, sociologists have already amassed a great deal of findings concerning inter-generational inequalities, one notable example being that it's very clear that large inequalities in parent's education tend to reproduce large inequalities in children's education \citep{breen2007}. Virtually in all countries, be it developed or developing, there is inequality of opportunity. But there is considerable variation in the magnitud of this effect. For example, the Scandinavian countries, particularly Denmark and Sweden, prove to be very mobile countries \citep{esping2012, breen2007, shavit1993}.

With very recent data, \citet{reardon_portilla} have unexpectedly uncovered a new finding: the reversal of the trend. From 1998 until 2010, it seems like the income achievement gap \footnote{The income achievement gap is the difference between the 90th and 10th income percentile in the test of interest} decreased at a rate of 0.01 and 0.008 standard deviations per year. In contrast to \citep{reardon2011}, we find that in a 30-year span the gap was systematicaly increasing at a rate of 0.02, something not dramatically different from the previous estimates. The American situation is a very distinctive one relative to Europe. The reasons why we're seeing a reversal in the trend could be numerous and should be studied very close to context. For example, the American educational system lacks any formal curricular differentiation \emph{à la} European style. In contrast, \citet{reardon2011} suggest that the reversal is likely due to the high increase of preschool enrollment. They suggest that in this same period (1998 - 2010) the income achievement gap in early schooling enrollment decreased substantially.

Some authors have taken this analysis to an international context in order to discover between-country trends. The comprehensive work of \citet{bradbury2015} performs a comparative analysis of Australia, United Kingdom, United States and Canada. Their research design is very distinctive in that they use longitudinal data from children as early as age 2 and study the evolution of the achievement gap up until age 14 \footnote{To the best of my knowledge this is the only study that uses panel data to study achievement gaps, let alone to do this between countries}. The core finding behind the book is that the American achievement gap is much wider than in any other comparison countries. As mentioned earlier, they find that once the achievement gap is present in early school entry, it doesn't seem to narrow very much over the life course. They find that the achievement gap is very stable all the way from early education to secondary school (although their studies cover up until age 14).

% You left documenting the bibliography here
One limitation of their study is that they concentrate on countries which have a very particular educational structure, namely the fact that there is little formal stratification in terms of curricula. These four countries have no major jump in school selection, quite the opposite to the European average countries. A thorough review by \citet{werfhorst_mijs} sheds some light on the subject. First and foremost, they gather substantive evidence showing that countries which have a highly tracked curriculum tend to have high levels of inequality, measured in terms of achievement gaps. Hanushek and Woessman (tracking study), use the TIMSS and PISA surveys to gauge whether highly tracked countries do indeed increase inequality after students pass the age at first selection of tracking. Moreover, Mickleweight \& Schnepf (2007) using similar data but different design finds that countries which have a high level of tracking, are distinctively unequal in the difference between the top 95th and bottom 5th performers. In fact, the difference in test scores between these two groups is about 10 times higher then the average annual gain of a year of schooling.

@@

This paper is interested in two questions. As explained in the literature review, we already know that U.S achievement gaps have been growing over time, and some international evidence suggests that it might be increasing in other countries as well. All of these studies pool all students together from different ages and estimate global trends.

We take a different approach and we want to study age-specific trends in achievement gaps. We're interested in studying a) the size of the achievement in a comparative perspective and its relationship to overall achievement levels, b) which countries are experiencing changes in the achievement gap, c) the rate at which each country-gap is widening/narrowing and d) establishing whether the gaps are widening/narrowing because particular groups are getting ahead/behind.

We develop each question separately for more detail.

a) We want to test the notion that better performing countries have lower leves of achievement gap. \citep{werfhorst_mijs} emphasize that there is empirical evidence that suggests this. This pattern is not so obvious. For example, countries with high leves of tracking could maximize student performance, specially the high SES students, raising their overall performance and thus raising the national performance score. But if the bottom performers are not gaining at the same rate, then the achievement gap will inevitable grow resulting in a high performing countries with widening achievement gap.

b) The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker that we though. After recording a SES gap increase of about 40\% in only 30 years, \citep{reardon_portilla} stress that they also found a significant decrease in only 15 years of data. This lends credibility that our estimations will show variation over time. As usual, their estimations are only for children going from 4th to 8th grade. We concentrate on age-specific groups, in this case, 15 year olds.

c) We want to compare the percentage change at which the gap widened/narrowed from the first to the last year. This will give us a general idea of the overall change over time, and will allow us to compare our estimates to the actual literature \footnote{Although no study has performed this age-specific achievement gap for comparable tests over such a long time. Our results will serve as comparison for other studies that use age-specific groups, such as 4th graders.}

d) The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment. The literature has concentrated very narrowly on whether the gap is increasing because the top performers are getting ahead, because both are distancing or because the bottom is falling behind. We shall pay particular attention to identifying the rate at which the top/bottom groups are evolving over time. This type of analysis is particularly useful in jotting down the mechanisms through which the gap is evolving. If the bottom performers are stagnated whereas the top is gaining ground, the efficiency-trade off dilemma of tracking might be much more credible than other explanations.

@@

The aim of this paper is to identify and disaggregate country trends in the achievement gap for several countries. To represent the SES gap, most of the literature on achievement gaps has concentrated on indicators such as parental education, parental occupational status, income achievement gaps and actual SES achievement gaps(Fryer and Levitt 2004). The actual calculation of the achievement gap varies substantially and different strategies have been implemented. For example, Mickleweight \& Funsche calculate the difference in achievement by crudely subtracting the gap between the 95th and 5th percentile of the mathematics distribution. Although in principle you should be able to capture some type of SES effect like this, theoretically, it should be much more accurate to difference out the mean score of, for example, parental education or some other SES proxy. Guan Kung Saw (2015), for instance, used parental education as a proxy of SES, whereas Soo-yong Byun and Kyung-keun Kim (2010) use a similar SES index as ours, but created by them.

\citet{reardon2011} and subsequently (all reardon papers), used a different method developed by \citet{reardon2011} which we partially adopt adopt in this paper. SES achievement gaps are measured as the difference in standardized achievement between the 90th and 10th percentiles of the chosen SES variable.

% Transform all of this text into equations

\[ x^n + y^n = z^n \]

\begin{math}
E=mc^2
\end{math}

We start by standardizing achievement for each country, year and subject. Because PISA does not provide a single achievement indicator, we take the median of all plausible values for each student \footnote{Since each plausible value is a random draw from a theoretical latent normal distribution of possible student achievement scores, the median should be precise in getting a central measure of the latent distribution.}, resulting in one single score. We then fit a linear model for each wave where we regress the single achievement score on the age variable weighted by the appropriate student sample weights. We take the residuals and divide them by the root mean square error of the model and divide that by one over the square root of the reliability of the survey for that year-subject (implement this all in formula). \footnote{Other proceedures multiply each country by their own reliability measure for each year-subject pair (Anna analysis of wordwide gap). Unfortunately, PISA 2000 did not provide any reliability measure separately for each country and PISA 2015 has to yet release their own at the moment of writing this paper. For this reason we implement the analysis following the original work of \citet{reardon2011}}

Standardizing the median test score solves the problem of comparability of gaps measured with different tests, and across waves. However, if the variance of academic achievement changes over time, then standardizing the overall score at each wave of testing confounds changes in the gap with changes in the variance of test scores. That is, by standardizing we're forcing the variance of test scores to be zero across all waves. But if the true variance of academic achievement grows over time, then the estimated trend in the achievement gaps will be underestimated, and vice versa.

% Show graph of variance over time
% I'm showing standard deviations, is it better to show variance?

Inspection of the data suggests that it's something we shouldn't be deeply concerned with. The standard deviation of each wave seems to be following a very similar pattern with an exception of the year 2000.

Another concern is that if the gaps at different waves are measured with tests that have different amounts of measurement error, then the amount of bias will not be the same in each measure of the gap. This can be very misleading and suggest erroneous intepretations regarding trends in the of the gaps over time \citet{reardon2011}. The PISA has tried to make sure the tests are comparable across waves (PISA 2012 technical report), but still we have to adjust for this imprecision.

In order to correct gap estimates for measurement error, I weight the student test scores by the P where P is the reliability of the test for each wave. Each PISA study provided a global reliability indicator which we use accordingly to each wave. This yields estimates of the true gaps, and eliminates any bias in the trend that may arise from differential reliability of the tests.

After constructing the adjusted test score measurement, we estimate the SES gap in a very simple fashion. For each country-wave, we estimate the weighted 90th and 10th percentile scores for the SES composite index, and then we generate a dummy of 1 for those above (including) the 90th percentile and 0 for those below (including) the 10th percentile. Finally, for each wave, we fit a mixed model with the adjusted test score as the dependent variable and the SES index as the only covariate. We allow both the intercept and the dummy variable to vary by country, and weight the model by the appropriate student weights.

We take the random effects of each country together with its standard error as the achievement gap. We fit a multilevel rather than a linear model because by pooling the information together, we weight countries appropriately to their sample size. Given that including the SES dummy reduces the sample size considerably, we want to be able to estimate each country-difference as accurately as possible.

% Show the properties of the ESCS dummy variable
% % in high educated
% % of home possessions
% % of ISEI

\section{Analysis}
